Variational Approximations for Transfer in Reinforcement Learning

### 2018-05-16 NIPS 2018/main.tex
Updated main paper with MGVT theorem instead of GVT

The second question that we need to answer is whether we can provide any guarantee on our algorithm's performance when given limited data. To address this point, we restrict to the case of linear approximators with Gaussian prior and posterior. We assume Alg. \ref{alg:transfer} is used with a finite dataset and provide a finite-sample analysis bounding the distance between the fixed-point of the mellow Bellman operator and $Q$-functions sampled from the variational distribution minimizing the objective \eqref{eq:elbo}. Our main results is given in the following theorem.

\begin{restatable}{theorem}{thg} \label{th:main-gaussian}
Fix a target task $\tau$ a let $\wt{Q}$ be the fixed-point of the corresponding mellow Bellman operator. Assume linearly parameterized value functions $Q_{\bm{w}}(s,a)=\bm{w}^T\bm{\phi}(s,a)$ with bounded weights $\bm{w}\leq w_{max}$ and uniformly bounded features $\bm{\phi}(s,a)\leq \phi_{\max}$. Consider the Gaussian version of Alg. \ref{alg:transfer} with prior $p(\bm{w}) =\mathcal{N}(\bm{\mu}_p,\bm{\Sigma}_p)$ and denote by $(\wh{\bm{\mu}},\wh{\bm{\Sigma}})$ the variational parameter minimizing the objective of Eq. \ref{eq:elbo} on a dataset $D$ of $N$ samples. Let $\nu$ be a probability measure over $\mathcal{S}\times\mathcal{A}$ and $\bm{w}^* = \arginf_{\bm{w}}\norm{\wt{B}(\bm{w})}_{\nu}^2$. Define $\upsilon(\bm{w}^*) \triangleq \mathbb{E}_{\mathcal{N}(\bm{w}^*,\bm{I})}\left[ v(\bm{w}) \right]$, with $v(\bm{w}) \triangleq \mathbb{E}_{\nu}\left[Var_{\mathcal{P}}\left[b(\bm{w})\right]\right]$. Then, there exist constants $c_1,c_2,c_3$ such that, with probability at least $1-2\delta$ over the choice of weights $\bm{w} \sim \mathcal{N}(\wh{\bm{\mu}},\wh{\bm{\Sigma}})$ and dataset $D$:
\begin{equation}
\resizebox{0.93\textwidth}{!}{$
\norm{Q_{\bm{w}} - \wt{Q}}_{\nu}^2 \leq \frac{1}{(1-\gamma)\delta}\left(2\norm{\wt{B}(\bm{w}^{*})}_{\nu}^2 + \frac{\upsilon(\bm{w}^*) + c_3\sqrt{\log\frac{2}{\delta}}}{\sqrt{N}} + \frac{c_2 + \lambda\norm{\bm{w}^* - \bm{\mu}_p}_{\bm{\Sigma}_p^{-1}}}{N} + \frac{c_1}{N^2}\right)$}
\end{equation}
\end{restatable}
We refer the reader to App. \ref{} for the proof and a specific definition of the three constants. Four main terms constitute our bound: the approximation error due to the limited hypothesis space (first term), the variance (second term), the distance to the prior (third term), and a constant term decaying very fast (as $\mathcal{O}(\frac{1}{N^2})$). Intuitively, our bound demonstrates that, if one starts very close the prior, the only error incurred, besides the irreducible approximation error, is due to the variance of the estimates, and, thus, the algorithm is expected to achieve good performance rather quickly. Furthermore, as $N\rightarrow\infty$ the only error term remaining is the irreducible approximation error. Notice that the variance term $\upsilon(\bm{w}^*)$ is due to the fact that we minimize a biased estimator of the Bellman error. As mentioned in the preliminaries, if double sampling of the next state is possible (e.g., in simulation), such term would disappear. Finally, we want to point out that the proof of this theorem relies on a very general result (Lemma \ref{lemma:general-bound} in App. \ref{}) that does not require any specific choice of distribution of approximator. Leveraging on such results, it is straightforward to provide a finite-sample analysis of the mixture version of our algorithm. For the sake of completeness, we report the derivation in App. \ref{}.

@@

\thg*
\begin{proof}
Using Lemma \ref{lemma:general-bound} with variational parameters $\wh{\bm{\xi}} = (\wh{\bm{\mu}},\wh{\bm{\Sigma}})$, we have:
\begin{align}\label{eq:thg-4}
\mathbb{E}_{q_{\wh{\bm{\xi}}}}\left[ \norm{B(\bm{w})}_{\nu}^2 \right ] \leq\ & \inf_{\bm{\xi} \in \Xi}\left\{ \mathbb{E}_{q_{\bm{\xi}}}\left[ \norm{B(\bm{w})}_{\nu}^2 \right ] + \mathbb{E}_{q_{\bm{\xi}}}\left[ v(\bm{w}) \right] + 2\frac{\lambda}{N} KL({q_{\bm{\xi}}}||p) \right\} + 2\frac{R_{max}^2}{(1-\gamma)^2}\sqrt{\frac{\log\frac{2}{\delta}}{N}}\notag\\ \leq\ & \mathbb{E}_{\mathcal{N}(\bm{w}^*,c\bm{I})}\left[ \norm{B(\bm{w})}_{\nu}^2 \right ] + \mathbb{E}_{\mathcal{N}(\bm{w}^*,c\bm{I})}\left[ v(\bm{w}) \right] + 2\frac{\lambda}{N} KL\left(\mathcal{N}(\bm{w}^*,c\bm{I})\ ||\ p\right)\notag\\ &\ \ + 2\frac{R_{max}^2}{(1-\gamma)^2}\sqrt{\frac{\log\frac{2}{\delta}}{N}}
\end{align}
where the second inequality is due to the fact that, since Lemma \ref{lemma:general-bound} contains an infimum over the variational parameters, we can upper bound its right-hand side by choosing any specific $\bm{\xi}$ from $\Xi$. Here, we choose $\bm{\mu} = \bm{w}^*$ and $\bm{\Sigma} = c\bm{I}$, for some positive constant $c>0$. Let us now bound these terms separately.

@@

\begin{restatable}{theorem}{thmg} \label{th:main-mixture}
Fix a target task $\tau$ a let $\wt{Q}$ be the fixed-point of the corresponding mellow Bellman operator. Assume linearly parameterized value functions $Q_{\bm{w}}(s,a)=\bm{w}^T\bm{\phi}(s,a)$ with bounded weights $\bm{w}\leq w_{max}$ and uniformly bounded features $\bm{\phi}(s,a)\leq \phi_{\max}$. Consider the mixture version of Alg. \ref{alg:transfer} using $C$ components, source task weights $\mathcal{W}_s$, and prior $p(\bm{w}) =\frac{1}{\abs{\mathcal{W}_s}}\sum_j\mathcal{N}(\bm{w}_j,\sigma^2_p\bm{I})$. Denote by $\wh{\bm{\xi}} = (\wh{\bm{\mu}}_1,\dots,\wh{\bm{\mu}}_C,\wh{\bm{\Sigma}}_1,\dots,\wh{\bm{\Sigma}}_C)$ the variational parameters minimizing the objective of Eq. \ref{eq:elbo} on a dataset $D$ of $N$ samples. Let $\nu$ be a probability measure over $\mathcal{S}\times\mathcal{A}$ and $\bm{w}^* = \arginf_{\bm{w}}\norm{\wt{B}(\bm{w})}_{\nu}^2$. Define $\upsilon(\bm{w}^*) \triangleq \mathbb{E}_{\mathcal{N}(\bm{w}^*,\bm{I})}\left[ v(\bm{w}) \right]$, with $v(\bm{w}) \triangleq \mathbb{E}_{\nu}\left[Var_{\mathcal{P}}\left[b(\bm{w})\right]\right]$. Then, there exist constants $c_1,c_2,c_3,c_4$ such that, with probability at least $1-2\delta$ over the choice of weights $\bm{w} \sim \frac{1}{C}\sum_i\mathcal{N}(\wh{\bm{\mu}}_i,\wh{\bm{\Sigma}}_i)$ and dataset $D$:
\begin{equation}
\resizebox{0.93\textwidth}{!}{$
\norm{Q_{\bm{w}} - \wt{Q}}_{\nu}^2 \leq \frac{1}{(1-\gamma)\delta}\left(2\norm{\wt{B}(\bm{w}^{*})}_{\nu}^2 + \frac{c_1}{N^2} + \frac{c_2 + \frac{c_4}{\abs{\mathcal{W}_s}}\sum_{j}\norm{\bm{w}^* - \bm{w}_{j}}}{N} + \frac{\upsilon(\bm{w}^*) + c_3\sqrt{\log\frac{2}{\delta}}}{\sqrt{N}}\right)$}
\end{equation}
\end{restatable}

>>>>

The second question that we need to answer is whether we can provide any guarantee on our algorithm's performance when given limited data. To address this point, we consider Alg. \ref{alg:transfer} with mixture of Gaussians and linear approximators. We assume Alg. \ref{alg:transfer} is used with a finite dataset and provide a finite-sample analysis bounding the distance between the fixed-point of the mellow Bellman operator and $Q$-functions sampled from the variational distribution minimizing the objective \eqref{eq:elbo}. Our main results is given in the following theorem.

\begin{restatable}{theorem}{thmg} \label{th:main-mixture}
Fix a target task $\tau$ a let $\wt{Q}$ be the fixed-point of the corresponding mellow Bellman operator. Assume linearly parameterized value functions $Q_{\bm{w}}(s,a)=\bm{w}^T\bm{\phi}(s,a)$ with bounded weights $\bm{w}\leq w_{max}$ and uniformly bounded features $\bm{\phi}(s,a)\leq \phi_{\max}$. Consider the mixture version of Alg. \ref{alg:transfer} using $C$ components, source task weights $\mathcal{W}_s$, and prior $p(\bm{w}) =\frac{1}{\abs{\mathcal{W}_s}}\sum_j\mathcal{N}(\bm{w}_j,\sigma^2_p\bm{I})$. Denote by $\wh{\bm{\xi}} = (\wh{\bm{\mu}}_1,\dots,\wh{\bm{\mu}}_C,\wh{\bm{\Sigma}}_1,\dots,\wh{\bm{\Sigma}}_C)$ the variational parameters minimizing the objective of Eq. \ref{eq:elbo} on a dataset $D$ of $N$ samples. Let $\nu$ be a probability measure over $\mathcal{S}\times\mathcal{A}$ and $\bm{w}^* = \arginf_{\bm{w}}\norm{\wt{B}(\bm{w})}_{\nu}^2$. Define $\upsilon(\bm{w}^*) \triangleq \mathbb{E}_{\mathcal{N}(\bm{w}^*,\bm{I})}\left[ v(\bm{w}) \right]$, with $v(\bm{w}) \triangleq \mathbb{E}_{\nu}\left[Var_{\mathcal{P}}\left[b(\bm{w})\right]\right]$. Then, there exist constants $c_1,c_2,c_3,c_4$ such that, with probability at least $1-2\delta$ over the choice of weights $\bm{w} \sim \frac{1}{C}\sum_i\mathcal{N}(\wh{\bm{\mu}}_i,\wh{\bm{\Sigma}}_i)$ and dataset $D$:
\begin{equation}
\resizebox{0.93\textwidth}{!}{$
\norm{Q_{\bm{w}} - \wt{Q}}_{\nu}^2 \leq \frac{1}{(1-\gamma)\delta}\left(2\norm{\wt{B}(\bm{w}^{*})}_{\nu}^2 + \frac{c_1}{N^2} + \frac{c_2 + \frac{c_4}{\abs{\mathcal{W}_s}}\sum_{j}\norm{\bm{w}^* - \bm{w}_{j}}}{N} + \frac{\upsilon(\bm{w}^*) + c_3\sqrt{\log\frac{2}{\delta}}}{\sqrt{N}}\right)$}
\end{equation}
\end{restatable}

We refer the reader to App. \ref{} for the proof and a specific definition of the three constants. Four main terms constitute our bound: the approximation error due to the limited hypothesis space (first term), the variance (second term), the distance to the prior (third term), and a constant term decaying as $\mathcal{O}(\frac{1}{N^2})$. Intuitively, the bound is tighten when the source tasks' $Q$-functions are, on average, close to the optimal ones for the target task. In such case, the dominating error is due to the variance of the estimates, and, thus, the algorithm is expected to achieve good performance rather quickly, as new data is collected. Furthermore, as $N\rightarrow\infty$ the only error term remaining is, as usual, the irreducible approximation error. Notice that the variance term $\upsilon(\bm{w}^*)$ is due to the fact that we minimize a biased estimator of the Bellman error. If double sampling of the next state were possible (e.g., in simulation), such term could be removed. Finally, we want to point out that the proof of this theorem relies on a very general result (Lemma \ref{lemma:general-bound} in App. \ref{}) that does not require any specific choice of distribution of approximator. Leveraging on such results, it is straightforward to provide a finite-sample analysis of the Gaussian version of our algorithm. For the sake of completeness, we report the derivation in App. \ref{}.\todo{Should we discuss the main difference between the MGVT and GVT bounds here?}

@@

\begin{restatable}{theorem}{thg} \label{th:main-gaussian}
Fix a target task $\tau$ a let $\wt{Q}$ be the fixed-point of the corresponding mellow Bellman operator. Assume linearly parameterized value functions $Q_{\bm{w}}(s,a)=\bm{w}^T\bm{\phi}(s,a)$ with bounded weights $\bm{w}\leq w_{max}$ and uniformly bounded features $\bm{\phi}(s,a)\leq \phi_{\max}$. Consider the Gaussian version of Alg. \ref{alg:transfer} with prior $p(\bm{w}) =\mathcal{N}(\bm{\mu}_p,\bm{\Sigma}_p)$ and denote by $(\wh{\bm{\mu}},\wh{\bm{\Sigma}})$ the variational parameter minimizing the objective of Eq. \ref{eq:elbo} on a dataset $D$ of $N$ samples. Let $\nu$ be a probability measure over $\mathcal{S}\times\mathcal{A}$ and $\bm{w}^* = \arginf_{\bm{w}}\norm{\wt{B}(\bm{w})}_{\nu}^2$. Define $\upsilon(\bm{w}^*) \triangleq \mathbb{E}_{\mathcal{N}(\bm{w}^*,\bm{I})}\left[ v(\bm{w}) \right]$, with $v(\bm{w}) \triangleq \mathbb{E}_{\nu}\left[Var_{\mathcal{P}}\left[b(\bm{w})\right]\right]$. Then, there exist constants $c_1,c_2,c_3$ such that, with probability at least $1-2\delta$ over the choice of weights $\bm{w} \sim \mathcal{N}(\wh{\bm{\mu}},\wh{\bm{\Sigma}})$ and dataset $D$:
\begin{equation}
\resizebox{0.93\textwidth}{!}{$
\norm{Q_{\bm{w}} - \wt{Q}}_{\nu}^2 \leq \frac{1}{(1-\gamma)\delta}\left(2\norm{\wt{B}(\bm{w}^{*})}_{\nu}^2 + \frac{\upsilon(\bm{w}^*) + c_3\sqrt{\log\frac{2}{\delta}}}{\sqrt{N}} + \frac{c_2 + \lambda\norm{\bm{w}^* - \bm{\mu}_p}_{\bm{\Sigma}_p^{-1}}}{N} + \frac{c_1}{N^2}\right)$}
\end{equation}
\end{restatable}
\begin{proof}
Using Lemma \ref{lemma:general-bound} with variational parameters $\wh{\bm{\xi}} = (\wh{\bm{\mu}},\wh{\bm{\Sigma}})$, we have:
\begin{align}\label{eq:thg-4}
\mathbb{E}_{q_{\wh{\bm{\xi}}}}\left[ \norm{B(\bm{w})}_{\nu}^2 \right ] \leq\ & \inf_{\bm{\xi} \in \Xi}\left\{ \mathbb{E}_{q_{\bm{\xi}}}\left[ \norm{B(\bm{w})}_{\nu}^2 \right ] + \mathbb{E}_{q_{\bm{\xi}}}\left[ v(\bm{w}) \right] + 2\frac{\lambda}{N} KL({q_{\bm{\xi}}}||p) \right\} + 2\frac{R_{max}^2}{(1-\gamma)^2}\sqrt{\frac{\log\frac{2}{\delta}}{N}}\notag\\ \leq\ & \mathbb{E}_{\mathcal{N}(\bm{w}^*,c\bm{I})}\left[ \norm{B(\bm{w})}_{\nu}^2 \right ] + \mathbb{E}_{\mathcal{N}(\bm{w}^*,c\bm{I})}\left[ v(\bm{w}) \right] + 2\frac{\lambda}{N} KL\left(\mathcal{N}(\bm{w}^*,c\bm{I})\ ||\ p\right)\notag\\ &\ \ + 2\frac{R_{max}^2}{(1-\gamma)^2}\sqrt{\frac{\log\frac{2}{\delta}}{N}}
\end{align}
where the second inequality is due to the fact that, since Lemma \ref{lemma:general-bound} contains an infimum over the variational parameters, we can upper bound its right-hand side by choosing any specific $\bm{\xi}$ from $\Xi$. Here, we choose $\bm{\mu} = \bm{w}^*$ and $\bm{\Sigma} = c\bm{I}$, for some positive constant $c>0$. Let us now bound these terms separately.

@@

\thmg*
